Using the internal language of toposes in algebraic geometry

### 2015-11-27 slides-ihes2015.tex
Final version of the slides

\newcommand{\backupstart}{
  \newcounter{framenumberpreappendix}
  \setcounter{framenumberpreappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
  \addtocounter{framenumber}{\value{framenumberpreappendix}} 
}

@@

\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}

\begin{frame}[plain,c]
  \centering
  \includegraphics[scale=0.37]{images/math-overflow-steven-gubkin}
  \par
\end{frame}

\section{The internal language}

@@

\begin{frame}\frametitle{Building a dictionary}
  \slogan{\hil{Understand notions of algebraic geometry
  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally & internally to $\Sh(X)$ \\
      \midrule
      sheaf of sets & set/type \\
      morphism of sheaves & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      \midrule
      sheaf of rings & ring \\
      sheaf of modules & module \\
      sheaf of finite type & finitely generated module \\
      locally free sheaf & free module \\
      tensor product of sheaves & tensor product of modules \\
      rank function & minimal number of generators \\
      sheaf of rational functions & total quotient ring of~$\O_X$ \\
      \bottomrule
    \end{tabular}
  \end{center}

@@

    \begin{minipage}{0.75\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
        of~$\O_X$-modules. If~$\F'$ and~$\F''$ are locally of finite type, so
        is~$\F$.
      \end{exampleblock}
    \end{minipage}
  \end{center}
\end{frame}

@@

    \begin{minipage}{0.70\textwidth}
      \begin{exampleblock}{}
        \justifying
        Any sheaf of modules of finite type on a reduced scheme is locally free
        \emph{on a dense open subset}.
      \end{exampleblock}
    \end{minipage}
  \end{center}
\end{frame}

@@

  \only<2->{
    Let~$X$ be a scheme. Depending on~$\Diamond$,~$\Sh(X)
    \models \Diamond\varphi$ means that~$\varphi$ holds on \ldots
    \begin{itemize}
      \item \ldots{} a dense open subset.
      \item \ldots{} a given open subset~$U$.
      \item \ldots{} an open subset containing a given closed subset~$A$.
      \item \ldots{} an open neighbourhood of a given point~$x \in X$.
    \end{itemize}

@@

\begin{frame}\frametitle{Quasicoherence}
  Let~$X$ be a scheme. Let~$\F$ be an~$\O_X$-module.

  Then~$\F$ is quasicoherent
  if and only if, internally to~$\Sh(X)$,
  \begin{quote}\textnormal{$\F[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
  \qquad\qquad where~$\Diamond_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
  \end{quote}
  \pause

  In particular: If~$\F$ is quasicoherent, then internally
  \[ (\text{$f$ invertible} \Rightarrow s = 0) \Longrightarrow
    \bigvee_{n \geq 0} f^n s = 0 \]
  \vspace*{-1.5em}\par%
  for any~$f : \O_X$ and~$s : \F$.
\end{frame}

@@

\newcommand{\defspeca}{\text{topological space of the prime ideals of $A$}}
\newcommand{\defspecb}{\text{topological space of the filters of $A$}}
\newcommand{\defspecc}{\text{locale of the filters of $A$}}

@@

\begin{frame}\frametitle{The relative spectrum}
  Let~$X$ be a scheme and~$\O_X \xrightarrow{\varphi} \A$ be a quasicoherent algebra.
  Can we describe~$\RelSpec_X \A$, a scheme over~$X$, internally?

@@

  Solution: Define internally the frame of~$\RelSpec_X \A$ to be the frame of
  those radical ideals~$I \subseteq \A$ such that
  \[ \forall f\?\O_X\_ \forall s\?\A\_
    (\text{$f$ invertible in~$\O_X$} \Rightarrow s \in I) \Longrightarrow
      fs \in I. \]
  \pause
  Its points are those filters~$\fff$ of~$\A$ such that
  \[ \forall f\?\O_X\_ \varphi(f) \in \fff \Rightarrow \text{$f$ invertible in~$\O_X$}. \]
\end{frame}

@@

  {\vspace{-0.5em}\centering
  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
  \includegraphics[scale=0.3]{images/external-internal-small}
  \par\medskip}

  \begin{itemize}
    \item Simplify proofs.
    \item Gain conceptual clarity.
    \item Develop a synthetic account of scheme theory.
  \end{itemize}
\end{frame}
% Fun facts:
% Take O_X, pullback along X_{negneg} --> X, pushforward: result is K_X.
% Very naive definition of P^n works internal to the big Zariski topos.

@@

@@

  \begin{itemize}
    \item \ldots{} $X$-schemes look just like sets,
    \item \ldots{} $\mathbb{P}^n_X$ is given by the naive expression
    \[ \{ (x_0,\ldots,x_n) \,|\, x_1 \neq 0 \vee \cdots \vee x_n \neq 0
    \}/\text{(rescaling)}, \]
    \item \ldots{} affinity is a ``double dual condition'', and
    \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ such
    that
    \[ \forall f : \mathbb{A}^1_X[T]\_
      \text{$f$ is separable} \Rightarrow
      \Diamond(\exists t : \mathbb{A}^1\_ f(t) = 0). \]
  \end{itemize}
\end{frame}

>>>>

\newcounter{framenumberpreappendix}
\newcommand{\backupstart}{
  \setcounter{framenumberpreappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
  \addtocounter{framenumber}{\value{framenumberpreappendix}} 
}

@@

\backupstart
\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}

\begin{frame}[plain,c]
  \centering
  \includegraphics[scale=0.37]{images/math-overflow-steven-gubkin}
  \par
\end{frame}
\backupend

\section{Basic applications of the internal language}

@@

\begin{frame}\frametitle{Building a dictionary}
  \slogan{\hil{Understand notions of algebraic geometry
  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
  \begin{center}
    \small
    \scalebox{0.93}{\begin{tabular}{ll}
      \toprule
      externally & internally to $\Sh(X)$ \\
      \midrule
      sheaf of sets & set/type \\
      morphism of sheaves & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      \midrule
      sheaf of rings & ring \\
      sheaf of modules & module \\
      sheaf of finite type & finitely generated module \\
      finite locally free sheaf & finite free module \\
      coherent sheaf & coherent module \\
      tensor product of sheaves & tensor product of modules \\
      rank function & minimal number of generators \\
      sheaf of rational functions & total quotient ring of~$\O_X$ \\
      \bottomrule
    \end{tabular}}
  \end{center}

@@

    \begin{minipage}{0.75\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
        of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type, so
        is~$\F$.
      \end{exampleblock}
    \end{minipage}
  \end{center}
\end{frame}

@@

    \begin{minipage}{0.70\textwidth}
      \begin{exampleblock}{}
        \justifying
        Any sheaf of modules of finite type on a reduced scheme is locally free
        \emph{on a dense open subset}.
      \end{exampleblock}
      \centering
      \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
      \par
    \end{minipage}
  \end{center}
\end{frame}

@@

  \only<2->{
    Let~$X$ be a scheme. Depending on~$\Diamond$,~$\Sh(X)
    \models \Diamond\varphi$ means that~$\varphi$ holds on \ldots
    \begin{itemize}
      \item \ldots{} a dense open subset.
      \item \ldots{} a schematically dense open subset.
      \item \ldots{} a given open subset~$U$.
      \item \ldots{} an open subset containing a given closed subset~$A$.
      \item \ldots{} an open neighbourhood of a given point~$x \in X$.
    \end{itemize}

@@

\begin{frame}\frametitle{Quasicoherence}
  Let~$X$ be a scheme. Let~$\E$ be an~$\O_X$-module.

  Then~$\E$ is quasicoherent
  if and only if, internally to~$\Sh(X)$,
  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
  \qquad\qquad where~$\Diamond_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
  \end{quote}
  \pause

  In particular: If~$\E$ is quasicoherent, then internally
  \[ (\text{$f$ invertible} \Rightarrow s = 0) \Longrightarrow
    \bigvee_{n \geq 0} f^n s = 0 \]
  \vspace*{-1.5em}\par%
  for any~$f : \O_X$ and~$s : \E$.
\end{frame}

@@

\newcommand{\defspeca}{\text{topological space of the prime ideals of $A$}}
\newcommand{\defspecb}{\text{topological space of the prime filters of $A$}}
\newcommand{\defspecc}{\text{locale of the prime filters of $A$}}

@@

\begin{frame}\frametitle{The relative spectrum}
  Let~$X$ be a scheme and~$\O_X \xrightarrow{\varphi} \A$ be a quasicoherent algebra.
  Can we describe~\hil{$\boldsymbol{\RelSpec_X \A}$}, a scheme over~$X$, internally?

@@

  \hil{Solution:} Define internally the frame of~$\RelSpec_X \A$ to be the frame of
  those radical ideals~$I \subseteq \A$ such that
  \[ \forall f\?\O_X\_ \forall s\?\A\_
    (\text{$f$ invertible in~$\O_X$} \Rightarrow s \in I) \Longrightarrow
      fs \in I. \]
  \pause
  Its \hil{points} are those prime filters~$G$ of~$\A$ such that
  \[ \forall f\?\O_X\_ \varphi(f) \in G \Rightarrow \text{$f$ invertible in~$\O_X$}. \]
\end{frame}

@@

  {\vspace{-0.5em}\centering
  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
  \includegraphics[scale=0.23]{images/external-internal-small}
  \par\medskip}

  \begin{itemize}
    \item Simplify proofs.
    \item Gain conceptual understanding.
    \item Develop a synthetic account of scheme theory.
  \end{itemize}

  \centering
  \hil{\href{http://tiny.cc/topos-notes}{http://tiny.cc/topos-notes}} \\
  \begin{minipage}{1.0\textwidth}\justifying
    \scriptsize
    spreading of properties,
    general transfer principles,
    applications to constructive algebra,
    quasicoherence,
    internal Cartier divisors,
    pullback along immersions $=$ internal sheafification,
    scheme dimension $=$ internal Krull dimension of~$\O_X$,
    dense $=$ not not,
    modal operators,
    relative spectrum,
    other toposes,
    étale topology,
    group schemes $=$ groups,
    \ldots
  \end{minipage}
  \par
\end{frame}
% Fun fact:
% Very naive definition of P^n works internal to the big Zariski topos.

@@

\begin{frame}\frametitle{Spreading from points to neighbourhoods}
  All of the following lemmas have a short, sometimes trivial proof.
  Let~$\F$ be a sheaf of finite type on a ringed space~$X$.
  Let~$x \in X$. Let~$A \subseteq X$ be a closed subset. Then:
  \small
  \begin{itemize}
    \item $\F_x = 0$ iff~$\F|_U = 0$ for some open neighbourhood of~$x$.
    \item $\F|_A = 0$ iff~$\F|_U = 0$ for some open set containing~$A$.
    \item $\F_x$ can be generated by~$n$ elements iff this is true on some open
    neighbourhood of~$x$.
%   \item $\alpha_x$ is surjective iff~$\alpha$ is an epimorphism on some open
%   set containing~$x$, where~$\alpha : \G \to \F$ is any morphism.
    \item $\mathcal{H}\mathrm{om}_{\O_X}(\F,\G)_x \cong
    \Hom_{\O_{X,x}}(\F_x,\G_x)$ if~$\F$ is of finite presentation around~$x$.
    \item $\F$ is torsion iff~$\F_\xi$ vanishes (assume~$X$ integral and~$\F$ quasicoherent).
    \item $\F$ is torsion iff~$\F|_{\mathrm{Ass}(\O_X)}$ vanishes (assume~$X$
    locally Noetherian and~$\F$ quasicoherent).
  \end{itemize}
\end{frame}

\begin{frame}\frametitle{The smallest dense sublocale}
  Let~$X$ be a reduced scheme satisfying a technical condition.
  Let~$i : X_{\neg\neg} \to X$ be the inclusion of the smallest dense sublocale
  of~$X$.

  Then~$i_* i^{-1} \O_X \cong \K_X$.

%  \begin{itemize}
%    \item This is a highbrow way of saying ``rational functions are regular
%    functions which are defined on a dense open subset''.
%    \item There is a generalization to nonreduced schemes.
%  \end{itemize}
\end{frame}

\begin{frame}\frametitle{Transfer principles}
  Let~$M$ be an~$A$-module. How do~$M$ and the sheaf~$M^{\sim}$ on~$\Spec A$
  relate?

  Observe that $M^{\sim} \cong \underline{M}[F^{-1}]$ is the localization of~$M$ at
  the \hil{generic filter}. Therefore:
  \begin{center}
    $M^{\sim}$ inherits all those properties of~$M$ which are \\
    \hil{stable under localization}.
  \end{center}
  Examples: finitely generated, free, flat, \ldots

  A converse holds as well, suitably formulated.
\end{frame}

@@

  \begin{itemize}
    \item \ldots{} $X$-schemes look just like sets,
    \item \ldots{} $\mathbb{P}^n_X$ is given by the naive expression
    \[ \{ (x_0,\ldots,x_n) \,|\, x_1 \neq 0 \vee \cdots \vee x_n \neq 0
    \}/\text{(rescaling)}, \]
    \item \ldots{} affinity is a ``double dual condition'', and
    \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ s.\,th.
    \[ \forall f : \mathbb{A}^1_X[T]\_\
      \text{$f$ is monic separable} \Rightarrow
      \Diamond(\exists t : \mathbb{A}^1\_ f(t) = 0). \]
  \end{itemize}
\end{frame}
